---
author: "Slava Nikitin"
date: "2017-02-13"
draft: false
tags: ["lecture"]
title: "Wrangle - Part I"
summary: "Tidy and Relational Data"
math: false
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
```{r echo = FALSE, warning=FALSE, message=FALSE}
library(readr)
library(tibble)
library(tidyr)
library(dplyr)
library(magrittr)
```


# Basic Concepts
## Material
- `tibble(x = 1:3)` local in R
- `data()` - local in R
- []("/Users/snikitin/Documents/training/class/analytics6600/static/data/movies.csv") - local on disc
- []("https://synikitin.github.io/analytics6600/data/movies.csv") - remote on disk, direct download
- []("https://www.quandl.com/tools/api") - remote on disc, api

## Machinery
- computer
- integrated development environment (IDE)
- programming language
- file formats

## Product
- data
- clean data
- visualization
- statistical model
- report
- web application
- optimized algorithm
- package

# Task
## IO
- `library(readr)`
- `library(tibble)`

## Tidy
- `library(tidyr)`

## Transform
- `library(dplyr)`

## Visualization
- `library(ggplot2)`

## Programming
- `library(magrittr)`

# Learning new packages and functions
- CRAN
- `help(package = "readr")`
- purpose
- available arguments
- example code, sometimes `demo(package = "stats")`, `example("summarise", package = "dplyr")`

# Tibbles  
## Creating
- `tibble`
tibble(
  population = c(783744, 923972, 823972, 789374),
  city = c("chicago", "columbus", "boston", "cleveland")
)

- `tribble`
tribble(
  ~x, ~y,
  1, "2",
  3, 4
)

## Printing
- `print(df, n, width)`
- `str`
- `View`

## Subsetting
- `df[["a"]]`, `df[[1]]`
- `df$a`

# Data Import
## read_csv
```
read_csv( 
  "Some random text here
  a,b,c
  1,2,! #This is a comment. ! is a missing value.
  4,5,6",
  skip = 2,
  na = "!",
  comment = "#",
  n_max = 2,
  col_names = c("x", "y", "z"),
  col_types = cols(x = col_integer(),
                   y = col_double(),
                   z = col_character())
)             
```

```{r eval = FALSE, echo = FALSE}
x <- "Confidential Data - Email abc@gmail.com for Permession
    $earnings,Months in 2017,store
    14343.43,January,! *Note: missing values marked with !
    34322.33,February,6343\n"
    
writeChar(  
  x,
  "readr_example.csv",
  176
)

read_csv(
  "readr_example.csv",
  skip = 2,
  na = "!",
  comment = "*",
  n_max = 2,
  col_names = c("earnings", "month2017", "store"),
  col_types = cols(store = col_character())
)

```

## arguments
- skip
- comment
- col_names
- na
- n_max
- locale
- col_types (logical, numbers, text, date, time, datetime)

## write_csv, write_excel_csv, write_rds
- similar arguments

## other packages
- haven
- readxl
- DBI
- jsonlite
- xml2

- Lets practice `read_csv` with the following data: 

```
x <- "Confidential Data - Email abc@gmail.com for Permission
    $earnings,Months in 2017,store
    14343.43,January,! *Note: missing values marked with !
    34322.33,February,6343\n"
    
writeChar(  
  x,
  "readr_example.csv",
  176
)

```


# Tidying Data
## Structure
- columns
- rows

## Semantics
- unit
- variable
- value
- observation

## Messy data
- Column headers are values, not variable names. This is messy
```{r }
a <- tribble(
  ~region, ~`$0 - $100k`, ~`$100k - $1,000k`, ~`$1,000k - $10,000k`,
  "midwest", 152, 78, 5,
  "west", 190, 130, 10
  )
a
```
and this is tidy by turning headers into a variable.
```{r}
gather(data = a, key = income, value = count, 2:4)
```


- Multiple variables are stored in one column. This is messy
```{r}
b <- tribble(
  ~`region-income`, ~count,
  "midwest, $0 - 100k", 152,
  "west, $0 - 100k", 190,
  "midwest, $100k - 1,000k", 78,
  "west, $100k - 1,000k", 130
)
b
```
and this is tidy by splitting a variable into two variables.
```{r}
separate(data = b, col = `region-income`,
         into = c("region", "income"), sep = ", ")
```



- Variables are stored in rows. This is messy
```{r}
d <- tribble(
  ~region, ~income, ~statistic,
  "midwest", 193478374, "annual_max",
  "midwest", 878743, "annual_min",
  "west", 29743443, "annual_max",
  "west", 19984028, "annual_min"
)
d
```
and this is tidy by turning a variable into column headers.
```{r}
spread(data = d, key = statistic, value = income)
```



- Multiple types of observational units are stored in the same table, in this case two separate time scales. This is messy
```{r}
e <- tribble(
  ~region, ~annual_max, ~annual_min, ~monthly_avg, ~month,
  "midwest", 193478374, 878743, 193947931, "January",
  "midwest", 193478374, 878743, 153394881, "February"
)
```
and this is tidy by removing annual scale variables
```{r}
select(e, -annual_max, -annual_min)
```
or this is tidy by removing monthly scale variables.
```{r}
select(e, -monthly_avg, -month) %>% 
  distinct
```


- A single observational unit is stored in multiple tables. This is messy
```{r}
f <- tribble(
  ~region, ~income, ~statistic,
  "midwest", 193478374, "annual_max",
  "midwest", 878743, "annual_min"
)
g <- tribble(
  ~region, ~income, ~statistic,
  "west", 29743443, "annual_max",
  "west", 19984028, "annual_min"
)
f
g
```
and this is tidy by combining them into a single data frame.
```{r}
bind_rows(f, g)
```



- Tidy the following dataset. It consists of a history of quarterly earnings per share of Johnson and Johnson. 
```{r}
df <- matrix(JohnsonJohnson, byrow = TRUE, ncol = 4) %>% 
  as_tibble() %>%
  rename(quarter1 = V1, quarter2 = V2, quarter3 = V3, quarter4 = V4) %>% 
  mutate(year = 1960:1980, stockinfo = "JNJ-NYSE")
df
```

```{r eval = FALSE, echo = FALSE}
gather(data = df, key = quarter, value = earnings, quarter1:quarter4) %>% 
  separate(col = stockinfo, into = c("symbol", "exchange"), sep = "-") #%>% 
  #select(exchange) %>% distinct
```

# Relational data
- tables and keys (primary, foreign, surrogate); key in the following data sets

```{r}
stocks <- tibble(earnings = c(800, 150, 240, 65),
                 volume = c(1599500, 30531626, 4276880, 25187460),
                 symbol = c("GOOG", "APPL", "GS", "MSFT"))
stocks
stocks2 <- tibble(earnings = c(800, 15),
                  volume = c(1599500, 13210000),
                  symbol = c("GOOG", "HBAN"))

symbols <- tibble(symbol = c("GOOG", "APPL", "GS", "HBAN"),
                  exchange = c("NASDAQ", "NASDAQ", "NYSE", "NASDAQ"))
symbols
```


- left, right, outer join (mutating)
```
inner_join(x = stocks, y = symbols, by = "symbol")
left_join(x = stocks, y = symbols, by = "symbol")
right_join(x = stocks, y = symbols, by = "symbol")
full_join(x = stocks, y = symbols, by = "symbol")
```

- semi, anti join (filtering)
```
semi_join(x = stocks, y = symbols, by = "symbol")
anti_join(x = stocks, y = symbols, by = "symbol")
```

- intersect, union, setdiff
```
union(x = stocks, y = stocks2)
intersect(x = stocks, y = stocks2)
setdiff(x = stocks, y = stocks2)
setdiff(x = stocks2, y = stocks)
```

- Identify primary and foreign keys in all datasets

```
set.seed(349834)

calls <- tibble(
  company = "schevchenko steele", #rep(companies, each = 12)
  number = rbinom(5, 30, .3),
  month = month.name[1:5] #rep(month.name, 10)
)
sales <- tibble(
  company = "schevchenko steele",
  amount = rgamma(10, 1e4, 1 / 1e4),
  month = month.name[-c(1, 9)]
)
orders <- tibble(
  company = "schevchenko steele",
  fullfilment = rbeta(10, 4, 8),
  month = month.name[-c(1, 9)]
) 
```
- Join calls and sales for only matching months
```{r echo = FALSE, eval = FALSE}
inner_join(calls, sales, c("company", "month"))
```

- Enrich calls with whatever sales data there is
```{r echo = FALSE, eval = FALSE}
left_join(calls, sales, c("company", "month"))
```

- Enrich sales with whatever calls data there is
```{r echo = FALSE, eval = FALSE}
right_join(calls, sales, c("company", "month"))
```

- Join calls and sales for all months
```{r echo = FALSE, eval = FALSE}
full_join(calls, sales, c("company", "month"))
```

- Obtain sales data for which you have calls
```{r echo = FALSE, eval = FALSE}
semi_join(sales, calls, c("company", "month"))
```

- Obtain orders data for which you have no calls 
```{r echo = FALSE, eval = FALSE}
anti_join(orders, calls, c("company", "month"))
```

- Join all datasets into a single data frame with all months
```{r echo = FALSE, eval = FALSE}
full_join(calls, sales, c("company", "month")) %>% 
  full_join(orders, c("company", "month"))
```






